\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{graphicx}
\graphicspath{{../../shared/figures/}}
\usepackage{natbib}
\bibliographystyle{unsrtnat}
\usepackage{doi}
% \setcitestyle{numbers,super} % Superscript citation numbers (e.g. ¬π, ¬≤‚Åª‚Å¥) ‚Äî does not support locator notes
\setcitestyle{numbers,square} % Square brackets around citation numbers (e.g. [1], [2-4]) ‚Äî math standard
% ‚îÄ‚îÄ Merge citation text into a single hyperlink ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ
\makeatletter
\renewcommand{\hyper@natlinkbreak}[2]{#1}
\makeatother
\usepackage{titlesec}
% New page before each top-level section
% \newcommand{\sectionbreak}{\clearpage}

% ‚îÄ‚îÄ Theorem-like environments ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{thmtools}
\declaretheorem[name=Theorem]{theorem}
\declaretheorem[name=Proposition,sibling=theorem]{proposition}
\declaretheorem[name=Lemma,sibling=theorem]{lemma}
\declaretheorem[name=Corollary,sibling=theorem]{corollary}
\declaretheorem[name=Conjecture,sibling=theorem]{conjecture}

\theoremstyle{definition}
\declaretheorem[name=Definition,style=definition]{definition}
\declaretheorem[name=Axiom,style=definition,sibling=definition]{axiom}
\declaretheorem[name=Assumption,style=definition,sibling=definition]{assumption}
\declaretheorem[name=Criterion,style=definition,sibling=definition]{criterion}
\declaretheorem[name=Example,style=definition,sibling=definition]{example}

\theoremstyle{remark}
\declaretheorem[name=Remark,style=remark]{remark}
\declaretheorem[name=Observation,style=remark,sibling=remark]{observation}
\declaretheorem[name=Property,style=remark,sibling=remark]{property}

% Prevent theorem-like environments from starting near the page bottom.
\usepackage{etoolbox}
\BeforeBeginEnvironment{theorem}{\needspace{4\baselineskip}\addvspace{6pt}}
\BeforeBeginEnvironment{corollary}{\needspace{4\baselineskip}\addvspace{6pt}}
\BeforeBeginEnvironment{lemma}{\needspace{4\baselineskip}\addvspace{6pt}}
\BeforeBeginEnvironment{proposition}{\needspace{4\baselineskip}\addvspace{6pt}}
\BeforeBeginEnvironment{definition}{\needspace{4\baselineskip}\addvspace{6pt}}
\BeforeBeginEnvironment{axiom}{\needspace{4\baselineskip}\addvspace{6pt}}
\BeforeBeginEnvironment{assumption}{\needspace{4\baselineskip}\addvspace{6pt}}
\BeforeBeginEnvironment{remark}{\needspace{4\baselineskip}\addvspace{9pt}}

% ‚îÄ‚îÄ Pandoc-generated table support ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{array}
\usepackage{calc}
\usepackage{caption}
% Pandoc 3.x wraps uncaptioned longtables in \def\LTcaptype{none};
% this requires a 'none' counter to exist.
\newcounter{none}

% ‚îÄ‚îÄ Paragraph spacing (matches Pandoc default) ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ
% No paragraph indentation; inter-paragraph spacing instead.
\usepackage{parskip}

% ‚îÄ‚îÄ Page-break quality ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ
% Prevent widow/orphan lines (single line stranded at top/bottom of page).
\widowpenalty=10000
\clubpenalty=10000
% Pin footnotes to the page bottom even with \raggedbottom.
\usepackage[bottom]{footmisc}
% Prevent page breaks immediately before or after display equations.
\predisplaypenalty=10000
\postdisplaypenalty=150
% Allow uneven page bottoms rather than stretching vertical glue.
\raggedbottom
% Increase separation between body text and footnotes.
\setlength{\skip\footins}{1.5\baselineskip}
% Ensure headings don't appear at page bottom without body text following.
\usepackage{needspace}
% Reserve enough space for the heading itself plus at least two following
% lines of body text (or a nested heading plus a line). This prevents a
% heading from appearing at the bottom of a page with its content (or its
% next-level subheading) starting on the following page.
\let\origsection\section
\renewcommand{\section}{\needspace{6\baselineskip}\origsection}
\let\origsubsection\subsection
\renewcommand{\subsection}{\needspace{6\baselineskip}\origsubsection}
\let\origsubsubsection\subsubsection
\renewcommand{\subsubsection}{\needspace{4\baselineskip}\origsubsubsection}
\let\origparagraphcmd\paragraph
\renewcommand{\paragraph}{\needspace{3\baselineskip}\origparagraphcmd}
\let\origsubparagraph\subparagraph
\renewcommand{\subparagraph}{\needspace{3\baselineskip}\origsubparagraph}

% Discourage page breaks inside multi-line math environments (align, gather, etc.)
\interdisplaylinepenalty=10000

% ‚îÄ‚îÄ Section numbering ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ
% Pandoc --number-sections generates \section{...} (not \section*{...}).
% LaTeX auto-numbers to depth 3 (subsubsection) by default.
\setcounter{secnumdepth}{3}

% ‚îÄ‚îÄ Make \paragraph and \subparagraph block-level (matches Pandoc) ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ
% By default LaTeX renders \paragraph as a run-in heading (text continues on
% the same line). These commands make them free-standing headings with
% proper spacing before and after.
\titleformat{\paragraph}
  {\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
\titlespacing*{\paragraph}
  {0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
\titleformat{\subparagraph}
  {\normalfont\normalsize\bfseries}{\thesubparagraph}{1em}{}
\titlespacing*{\subparagraph}
  {0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}

% ‚îÄ‚îÄ Pandoc-generated tightlist support ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% ‚îÄ‚îÄ Pandoc 3.x figure sizing ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ
% \pandocbounded constrains included graphics to text width when they exceed it
\newcommand{\pandocbounded}[1]{%
  \sbox0{#1}%
  \ifdim\wd0>\linewidth
    \resizebox{\linewidth}{!}{#1}%
  \else
    #1%
  \fi
}

% ‚îÄ‚îÄ Unicode character fixes ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ
\usepackage{amssymb}
\usepackage{newunicodechar}
\newunicodechar{‚úì}{$\checkmark$}
\newunicodechar{‚âÝ}{$\neq$}
\newunicodechar{‚â°}{$\equiv$}
\newunicodechar{‚âà}{$\approx$}
\newunicodechar{Œ¥}{$\delta$}
\newunicodechar{‚àí}{$-$}
% Superscript digits used in scientific notation (e.g. 10^-7, 10^29)
\newunicodechar{‚Å∞}{$^{0}$}
% ¬π¬≤¬≥ already defined by textcomp; suppress redefinition warning
\begingroup\renewcommand\PackageWarning[2]{}%
\newunicodechar{¬π}{$^{1}$}
\newunicodechar{¬≤}{$^{2}$}
\newunicodechar{¬≥}{$^{3}$}
\endgroup
\newunicodechar{‚Å¥}{$^{4}$}
\newunicodechar{‚Åµ}{$^{5}$}
\newunicodechar{‚Å∂}{$^{6}$}
\newunicodechar{‚Å∑}{$^{7}$}
\newunicodechar{‚Å∏}{$^{8}$}
\newunicodechar{‚Åπ}{$^{9}$}
\newunicodechar{‚Åª}{$^{-}$}
% Fix \rightarrow for use outside math mode
\let\mystRightarrow\rightarrow
\renewcommand{\rightarrow}{\ensuremath{\mystRightarrow}}
% ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ‚îÄ

% Allow TeX extra stretch to resolve moderate overfull hboxes
\emergencystretch=2em

\hypersetup{colorlinks=true, allcolors=blue}

\title{Thermodynamics of Cooperation: Cooperative Equilibrium}
\author{Keith Lostracco}
\date{}

\begin{document}
\maketitle

\begin{abstract}
We prove that in a repeated multi-agent game with energy-denominated
payoffs derived from thermodynamic first principles, universal
cooperation is the unique Nash equilibrium that simultaneously satisfies
Pareto efficiency and welfare maximization. The payoff matrix is not
stipulated but constructed from physical parameters (resource energy,
friction multipliers, exploitation efficiency, and repair costs),
yielding a Prisoner's Dilemma in which mutual defection is net-negative
(\(P < 0\)) rather than merely suboptimal. This structural property
drives the cooperation threshold to \(\delta^* = 0.363\), substantially
below the canonical \(0.5\). We prove that defection is evolutionarily
unstable, self-extinguishing in friction-coupled networks, and
unprofitable even from a single deviation when network effects are
included. The friction state variable, encoding the cumulative memory of
past violations, enters payoffs endogenously, making punishment
self-reinforcing without external enforcement. We explicitly
characterize the relationship between these results and established
repeated-game theory: the known machinery is the Folk Theorem and ESS
analysis; the novel content is the derived payoff matrix, structurally
negative mutual defection, equilibrium selection, and endogenous
persistent friction.
\end{abstract}
\newpage
\tableofcontents
\newpage
\section{Introduction}\label{introduction}

The classical repeated Prisoner's Dilemma demonstrates that cooperation
can be sustained as a Nash equilibrium when agents are sufficiently
patient \citep{Fudenberg1986, Fudenberg1991}. The Folk Theorem
establishes this as a general principle: any feasible, individually
rational payoff vector is achievable in equilibrium for sufficiently
high discount factors. This generality, while powerful, renders the
theorem indeterminate: it sustains cooperation but also sustains
exploitation, with no internal mechanism to select among the continuum
of equilibria.

This paper constructs a repeated game in which the payoff matrix is not
stipulated but derived from physical parameters. \emph{Necessary
Constraints} \citep{TC-I}, \emph{Thermodynamic Friction} \citep{TC-IV},
and \emph{Strategic Entropy Injection} \citep{TC-II} establish the
physical infrastructure: boundary constraints as Lagrangian multipliers
on multi-agent optimization, thermodynamic friction as the energy cost
of constraint violation, and information-theoretic costs of strategic
signal corruption. We inherit those results and construct the two-player
and \(N\)-player games with payoffs denominated in energy units
(Joules).

The energy denomination imposes structural constraints on the admissible
payoff space that break the Folk Theorem's indeterminacy. In particular:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The payoff matrix is \emph{derived} rather than assumed, with the
  Prisoner's Dilemma ordering emerging as a theorem
  (Proposition~\ref{prop-physical-prisoners-dilemma}).
\item
  Mutual defection is net-negative (\(P < 0\)) rather than merely
  suboptimal, because thermodynamic friction makes conflict
  self-destructive.
\item
  The resulting cooperation threshold \(\delta^* = 0.363\) is
  substantially below the canonical \(0.5\).
\item
  A friction state variable encoding the cumulative memory of past
  violations enters payoffs endogenously, making punishment
  self-reinforcing.
\end{enumerate}

All formal results are stated as numbered theorems with full proofs.

The paper proceeds as follows: model and physical setup,
energy-denominated payoff model, two-player Prisoner's Dilemma ordering,
cooperation theorem in the repeated game, network friction effects,
\(N\)-player public goods extension, evolutionary stability, tit-for-tat
dynamics, welfare uniqueness, and discussion.

\subsection{Related Work}\label{related-work}

The game-theoretic results build on two established traditions:
classical and evolutionary game theory, and the theory of repeated
games. Our contribution is the energy-denominated payoff matrix derived
from the dissipative-system model of \citep{TC-I, TC-IV}, together with
the cooperation, stability, and welfare consequences that follow from
its structural properties.

\textbf{Classical and evolutionary game theory.} \citet{VonNeumann1944}
created the mathematics of strategic interaction. \citeauthor{Nash1950}
\citetext{\citeyear{Nash1950}; \citeyear{Nash1951}} proved that every
finite game has at least one equilibrium in mixed strategies.
\citet{MaynardSmith1973} fused game theory with Darwinian selection,
introducing the Evolutionarily Stable Strategy (ESS), a concept strictly
stronger than Nash equilibrium that identifies strategies resistant to
invasion by mutants. \citet{Hamilton1964a} explained altruism through
inclusive fitness (\(rB > C\)). \citet{Axelrod1984} demonstrated that
Tit-for-Tat wins iterated Prisoner's Dilemma tournaments, establishing
that cooperation can emerge among self-interested agents without central
authority. \citet{Taylor1978} and \citet{Weibull1995} developed
replicator dynamics connecting game-theoretic equilibria to evolutionary
population dynamics. This literature uses dimensionless utility
(``utils''), a quantity that does not correspond to any physical
observable and is not interpersonally comparable. Our framework replaces
utils with energy units, inheriting the physical constraints that energy
must satisfy (conservation, positivity of costs, irreversibility of
entropy increase).

\textbf{The Folk Theorem and equilibrium selection.}
\citet{Fudenberg1986} proved the definitive Folk Theorem: when agents
are sufficiently patient, any feasible and individually rational outcome
can be sustained as a subgame-perfect Nash equilibrium. This result
simultaneously supports cooperation and creates indeterminacy: the
theory predicts everything and therefore nothing.
\citet{OsborneRubinstein1994} provide a comprehensive treatment of
equilibrium concepts in repeated games. Our framework resolves the Folk
Theorem's indeterminacy through the structural properties of the
energy-denominated payoff matrix: the net-negative mutual defection
payoff (\(P < 0\)) and the endogenous friction state variable together
select universal cooperation as the unique efficient equilibrium.

\textbf{Energy-denominated and physically grounded payoffs.} The
ecological economics tradition \citep{Lotka1922, GeorgescuRoegen1971}
recognized that energy is the fundamental currency of biological and
economic activity, but did not build individual-level game theory on
that foundation. \citet{Hirshleifer1991} modeled conflict as
resource-consuming economic activity but used abstract utility. The
thermodynamic friction framework \citep{TC-IV} provides the missing
link: contest dissipation, boundary damage, and repair costs yield a
friction multiplier \(\Phi\) that denominates the Prisoner's Dilemma
payoff matrix entirely in energy units. The resulting payoff structure
(\(P < 0\) under realistic parameters) is derived rather than
stipulated, and its consequences for equilibrium selection, evolutionary
stability, and welfare uniqueness are the central contributions of this
work.

\section{Model and Physical Setup}\label{model-and-physical-setup}

\subsection{Energy-Denominated
Payoffs}\label{energy-denominated-payoffs}

We adopt the physical framework established in \emph{Necessary
Constraints} \citep{TC-I} and \emph{Thermodynamic Friction}
\citep{TC-IV}, summarized here to establish notation. Two agents contest
a resource of energy value \(E_R > 0\). Each selects from two
meta-strategies:

\begin{itemize}
\tightlist
\item
  \textbf{Cooperate (C):} Accept the constrained allocation (the
  variational equilibrium of \citep{TC-I}). Pay only the coordination
  cost \(c \ll E_R\).
\item
  \textbf{Defect (D):} Violate the constraint. Seize the resource by
  force or strategic entropy injection, triggering the conflict contest
  dynamics of \citep{TC-IV} and/or the communication corruption dynamics
  of \citep{TC-II}.
\end{itemize}

The exploitation efficiency \(\theta \in (0,1)\), damage fractions
\(\delta^{\text{off}}, \delta^{\text{def}} \in [0,1]\), repair
multiplier \(\kappa \geq 1\), and friction multiplier
\(\Phi = 1 + (1+\kappa)(\delta^{\text{off}} + \delta^{\text{def}})\)
retain their definitions from \citep{TC-IV}. The bare \(\delta\) (no
superscript) denotes the discount factor throughout; context
distinguishes it from the superscripted damage fractions.

\subsection{The Friction State
Variable}\label{the-friction-state-variable}

Defection does not merely affect the immediate interaction; it injects
friction into the network, degrading all future cooperative
transactions. The friction dynamics, formalized in \citep{TC-IV},
involve a network friction coefficient \(\phi \geq 0\) that
parameterizes per-transaction overhead costs due to monitoring,
verification, and precautionary defense. Each boundary violation of
severity \(v\) injects \(\Delta\phi = \eta v\) into the friction
coefficient (where \(\eta > 0\) is the friction sensitivity), while in
the absence of new violations, friction decays at rate
\(\epsilon \in (0,1)\) per period.

The combined law of motion is:

\[\phi_{t+1} = (1 - \epsilon)\,\phi_t + \eta\,v_t \cdot \mathbb{1}_{\{\text{violation at } t\}}\]

The cascading cost of a single violation amplifies at ratios of \(10^3\)
to \(10^4\) across the cooperative network \citep{TC-IV}: a violation of
severity \(v\) produces total cascading cost
\(C_{\text{cascade}} = \eta v M \bar{E} / \epsilon\) over the full
recovery period.

The friction state variable enters our analysis through the
network-adjusted payoffs
(¬ß\ref{network-effects-the-friction-amplifier}). The key property is
persistence: violations today degrade payoffs for approximately
\(1/\epsilon\) future periods, coupling the repeated game to a dynamic
system absent from standard models.

\section{The Energy-Denominated Payoff
Model}\label{the-energy-denominated-payoff-model}

\subsection{Motivation: Why Energy Changes
Everything}\label{motivation-why-energy-changes-everything}

Standard game theory uses abstract utility values (often called
``points'' or ``utils''). The Prisoner's Dilemma, for example, assigns
payoffs like (3, 3) for mutual cooperation and (1, 1) for mutual
defection, without specifying what these numbers represent or where they
come from.

Our framework replaces abstract utils with \textbf{thermodynamic energy}
(Joules, Calories, or normalized energy units). This is not merely a
relabeling: it imports the physical constraints that energy must
satisfy:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Conservation:} Energy is neither created nor destroyed.
  Conflict does not generate new energy; it dissipates existing energy
  as waste heat \citep{TC-IV}.
\item
  \textbf{Positivity of costs:} All physical actions require energy
  expenditure (\(W = \int F \cdot dx \geq 0\)). Conflict investments,
  boundary damage, and repair are non-negative real costs, physically
  determined rather than stipulated.
\item
  \textbf{Diminishing returns:} The concavity of utility in resources
  (the regularity assumption Assumption 1 \citep{TC-I}, part (b)) means
  that marginal gains decrease while marginal losses increase, a
  thermodynamic asymmetry that breaks the symmetry of abstract payoffs.
\item
  \textbf{Network coupling:} An agent's payoff depends not only on the
  current interaction but on the network friction state \(\phi\), which
  persists across interactions \citep{TC-IV}.
\end{enumerate}

These physical constraints \textbf{structurally determine} the payoff
matrices. The cooperative equilibrium is not an assumption fed into the
model; it is a \textbf{consequence} of the physics.

\subsection{Strategy Space}\label{strategy-space}

Each agent selects from two meta-strategies in any pairwise interaction
over a contested resource:

\begin{definition}[Cooperative and Defection Strategies]
\label{def-cooperative-defection-strategies}

Each agent selects one of two meta-strategies per interaction:

\begin{itemize}
\item \textbf{Cooperate (C):} Respect the boundary constraint (Definition 3 \citep{TC-I}). Accept the constrained allocation $\mathbf{x}_i^*$ determined by the variational equilibrium (Theorem 7 \citep{TC-I}). Incur only the trade/coordination cost $c > 0$ (the energy cost of negotiation, contract enforcement, or social coordination).
\item \textbf{Defect (D):} Violate the boundary constraint. Attempt to seize the resource by force or deception. This triggers the conflict contest (Definition 3 \citep{TC-IV}) and/or deception channel (Definition 3 \citep{TC-II}), incurring the associated costs.
\end{itemize}

\end{definition}

\textbf{Interpretation:} ``Cooperate'' means operating within the
constraint system. ``Defect'' means violating constraints. The
distinction is energetic: whether to expend resources on coordination or
on conflict.

\subsection{The Energy Variables}\label{the-energy-variables}

For a pairwise interaction over resource \(E_R > 0\):

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Symbol
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Meaning
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\(E_R\) & Energy value of the contested resource & Physical
environment \\
\(c\) & Coordination cost (trade negotiation, contract) & Cooperative
overhead; \(c \ll E_R\) \\
\(e^*\) & Nash equilibrium contest investment per agent & Proposition 1
\citep{TC-IV}: \(e^* = E_R/4\) (symmetric) \\
\(\Phi\) & Total friction multiplier & \citep{TC-IV}:
\(\Phi = 1 + (1+\kappa)(\delta^{\text{off}} + \delta^{\text{def}})\) \\
\(\phi_0\) & Current network friction coefficient & Definition 7
\citep{TC-IV} \\
\(\Delta\phi\) & Friction injection from defection & Definition 8
\citep{TC-IV}: \(\Delta\phi = \eta \cdot v\) \\
\(\bar{E}\) & Average cooperative transaction value & Network
parameter \\
\(M\) & Number of transactions per period & Network parameter \\
\end{longtable}
}

\subsection{Admissibility: What Energy Denomination
Forbids}\label{admissibility-what-energy-denomination-forbids}

\begin{remark}[Admissibility of Energy-Denominated Payoff Matrices]
\label{rmk-admissibility}

The energy denomination restricts the class of admissible payoff matrices to those satisfying three constraints:

(a) \textbf{Resource-bounded payoffs.} All payoffs are bounded by the resource endowment: $|T|, |R|, |P|, |S| \leq E_R$. No player can extract more energy than the resource contains.

(b) \textbf{Strictly positive conflict costs.} Offensive and defensive actions require $e^* > 0$ energy investment. Costless aggression is physically inadmissible.

(c) \textbf{Persistent state coupling.} Payoffs in period $t$ depend on the network friction state $\phi_t$, which is a function of all prior interactions. Payoffs are not independently and identically distributed across periods.

Any payoff matrix violating (a)‚Äì(c) is physically inadmissible under energy denomination. Constraints (a)‚Äì(b) restrict the payoff space; constraint (c) restricts the game form.

A further property is not assumed but \textbf{derived} from (a)‚Äì(b) combined with the friction multiplier (Theorem 6 \citep{TC-IV}):

\textbf{Derived consequence: Net-negative mutual defection.} When $\Phi > 2$ (which holds for all physically realistic parameter ranges), the mutual defection payoff satisfies $P < 0$. Both agents lose energy in absolute terms. This is a structural consequence of the friction multiplier: no parameter is set to produce it. The derivation chain is: boundary damage exists on both sides ($\delta^{\text{off}} + \delta^{\text{def}} > 0$, since costless conflict is excluded by (b)); repair is thermodynamically irreversible since repair requires thermodynamic work exceeding the damage energy ($\kappa > 1$, by the Second Law); therefore $(1+\kappa)(\delta^{\text{off}} + \delta^{\text{def}}) > 1$, giving $\Phi > 2$ and $P = \frac{E_R}{2}(1 - \Phi/2) < 0$. See ¬ß\ref{construction} for the full derivation.

In particular, the standard Prisoner's Dilemma with $P > 0$ is the low-friction limit of this model, arising only when $\Phi < 2$, i.e., $(1+\kappa)(\delta^{\text{off}} + \delta^{\text{def}}) < 1$. At $\kappa = 2$ this requires combined damage $< 1/3$, the mild end of the conflict spectrum (low-stakes disputes, minor skirmishes). Once damage and repair costs are even moderately realistic, $\Phi > 2$ and $P < 0$.

\textbf{What is excluded:} Payoff specifications with unbounded gains, costless aggression, or memoryless (i.i.d.) interaction histories are inadmissible. Positive mutual-defection payoffs ($P > 0$) are not excluded by fiat but are \emph{proved impossible} under realistic friction (derived consequence above). These exclusions are non-trivial: they eliminate entire families of payoff matrices that are routinely studied in abstract game theory but cannot arise when payoffs are denominated in conserved physical quantities subject to thermodynamic friction.
\end{remark}

\section{The Two-Player Payoff
Matrix}\label{the-two-player-payoff-matrix}

\subsection{Construction}\label{construction}

Consider agents \(A\) and \(B\) interacting over resource \(E_R\). Using
the energy variables defined above, with the equal-split convention
under cooperation:

\textbf{Case (C, C): Mutual Cooperation:}

Both agents respect the constraint. The resource is split according to
the variational equilibrium (equal split for symmetric agents). Each
pays the coordination cost \(c\).

\[\pi_A^{CC} = \pi_B^{CC} = \frac{E_R}{2} - c\]

\textbf{Case (D, C): Unilateral Defection (A defects, B cooperates):}

Agent \(A\) attacks the cooperating agent \(B\). Since \(B\) is not
investing in contest (expecting cooperation), \(A\) seizes the resource
but still incurs: - The metabolic cost of the aggressive action:
\(e_A^{\text{attack}}\) (energy of mounting the violation) - Boundary
damage to self (offensive):
\(\delta^{\text{off}} \cdot e_A^{\text{attack}}\) (exposure cost) -
Repair of self-damage:
\(\kappa \cdot \delta^{\text{off}} \cdot e_A^{\text{attack}}\)

Agent \(B\), caught unprepared, suffers: - Loss of the full resource:
\(-E_R\) (relative to the cooperative baseline) - Defensive boundary
damage: \(\delta^{\text{def}} \cdot e_A^{\text{attack}}\) - Repair cost:
\(\kappa \cdot \delta^{\text{def}} \cdot e_A^{\text{attack}}\)

However, the defector against a cooperator faces much lower resistance.
We model the attack cost against an undefended boundary as
\(e_A^{\text{attack}} = \theta \cdot E_R\) where \(\theta \in (0, 1)\)
is the \textbf{exploitation efficiency}, the fraction of the resource's
value the defector must spend to seize it from an unprepared cooperator.
Typically \(\theta \ll 1/2\) (attacking an undefended target is cheap).

\[\pi_A^{DC} = E_R - \theta E_R \left[1 + (1 + \kappa)\delta^{\text{off}}\right] = E_R\left[1 - \theta(1 + (1+\kappa)\delta^{\text{off}})\right]\]

\[\pi_B^{DC} = -\theta E_R (1 + \kappa)\delta^{\text{def}}\]

The victim \(B\) receives nothing from the resource and additionally
suffers boundary damage and repair costs from the attack.

\textbf{Case (C, D): Symmetric mirror of (D, C):}

\[\pi_A^{CD} = -\theta E_R (1 + \kappa)\delta^{\text{def}}\]

\[\pi_B^{CD} = E_R\left[1 - \theta(1 + (1+\kappa)\delta^{\text{off}})\right]\]

\textbf{Case (D, D): Mutual Defection:}

Both agents fight. This is the Tullock contest from the thermodynamic
friction derivation (Proposition 1 \citep{TC-IV}). Each invests
\(e^* = E_R/4\), wins with probability 1/2, and suffers boundary damage
from both offense and defense plus repair costs.

Each agent's expected payoff is the winning prize minus all costs:

\[\pi_A^{DD} = \pi_B^{DD} = \frac{E_R}{2} - e^* - (\delta^{\text{off}} + \delta^{\text{def}}) e^* - \kappa(\delta^{\text{off}} + \delta^{\text{def}}) e^*\]

\[= \frac{E_R}{2} - \left[1 + (1 + \kappa)(\delta^{\text{off}} + \delta^{\text{def}})\right] e^* = \frac{E_R}{2} - \Phi \cdot \frac{E_R}{4}\]

\[= \frac{E_R}{2}\left(1 - \frac{\Phi}{2}\right)\]

\subsection{The Payoff Matrix
(Single-Round)}\label{the-payoff-matrix-single-round}

Assembling into standard normal form, define the shorthand payoffs:

\[T = \pi^{DC}_{\text{defector}} = E_R\left[1 - \theta(1 + (1+\kappa)\delta^{\text{off}})\right] \quad \text{(Temptation)}\]

\[R = \pi^{CC} = \frac{E_R}{2} - c \quad \text{(Reward for cooperation)}\]

\[P = \pi^{DD} = \frac{E_R}{2}\left(1 - \frac{\Phi}{2}\right) \quad \text{(Punishment for mutual defection)}\]

\[S = \pi^{CD}_{\text{sucker}} = -\theta E_R (1+\kappa)\delta^{\text{def}} \quad \text{(Sucker's payoff)}\]

The normal-form game with row player \(A\) and column player \(B\) is
then:

\[\begin{array}{c|c|c}
 & B: C & B: D \\ \hline
A: C & (R,\; R) & (S,\; T) \\ \hline
A: D & (T,\; S) & (P,\; P) \\
\end{array}\]

\subsection{Parameterization: The Physical Prisoner's
Dilemma}\label{parameterization-the-physical-prisoners-dilemma}

The classic Prisoner's Dilemma requires \(T > R > P > S\). Let us verify
each inequality under physical constraints:

\textbf{Claim: \(T > R\) (temptation exceeds reward) when exploitation
is cheap.}

\[T > R \iff E_R[1 - \theta(1 + (1+\kappa)\delta^{\text{off}})] > \frac{E_R}{2} - c\]

\[\iff \frac{E_R}{2} + c > \theta E_R(1 + (1+\kappa)\delta^{\text{off}})\]

\[\iff \theta < \frac{1/2 + c/E_R}{1 + (1+\kappa)\delta^{\text{off}}}\]

For \(c \ll E_R\) and \(\delta^{\text{off}} = 0.15\), \(\kappa = 2\):
threshold \(\approx 0.5/1.45 \approx 0.345\). So \(T > R\) when the
exploitation cost \(\theta < 0.345\), i.e., when stealing from a
cooperator is less than 34.5\% of the resource value. This is the
standard condition: defection is \emph{tempting} when exploiting a
cooperator is cheap.

\textbf{Claim: \(R > P\) (cooperation beats mutual defection).}

\[R > P \iff \frac{E_R}{2} - c > \frac{E_R}{2}\left(1 - \frac{\Phi}{2}\right) = \frac{E_R}{2} - \frac{\Phi E_R}{4}\]

\[\iff \frac{\Phi E_R}{4} > c\]

Since \(\Phi \geq 1\) and \(c \ll E_R\), this holds for any non-trivial
friction multiplier. Even with zero boundary damage (\(\Phi = 1\)),
mutual cooperation beats mutual defection as long as the coordination
cost is less than \(E_R/4\), i.e., less than the per-agent contest
dissipation.

\textbf{Claim: \(P > S\) (mutual defection beats being the sucker) when
friction is not too extreme.}

\[P > S \iff \frac{E_R}{2}\left(1 - \frac{\Phi}{2}\right) > -\theta E_R(1+\kappa)\delta^{\text{def}}\]

\[\iff \frac{1}{2} - \frac{\Phi}{4} + \theta(1+\kappa)\delta^{\text{def}} > 0 \iff \Phi < 2 + 4\theta(1+\kappa)\delta^{\text{def}} \equiv \Phi_{\text{PD}}\]

For the baseline parameters (\(\theta = 0.15\), \(\kappa = 2\),
\(\delta^{\text{def}} = 0.25\)):
\(\Phi_{\text{PD}} = 2 + 4(0.15)(3)(0.25) = 2.45\). Since the baseline
\(\Phi = 2.2 < 2.45\), the PD ordering \(P > S\) holds. For
\(\Phi \geq \Phi_{\text{PD}}\), the game transitions to a
\textbf{Chicken} (Hawk-Dove) structure with \(T > R > S > P\): mutual
defection becomes \emph{worse} than being the sucker, because the
contest dissipation at high friction overwhelms the sucker's one-sided
damage. This transition is \emph{favorable} for cooperation: the
repeated-game cooperation condition
(Theorem~\ref{thm-cooperation-theorem}) becomes easier to satisfy since
\(T - P\) grows larger, and even the one-shot game admits cooperative
mixed-strategy Nash equilibria under the Chicken structure.

\textbf{Claim: \(S < 0\) (the sucker always loses).}

\[S = -\theta E_R(1 + \kappa)\delta^{\text{def}} < 0\]

since all parameters are positive. The sucker not only fails to obtain
the resource but suffers boundary damage from the defector's attack.

\medskip

\begin{proposition}[Physical Prisoner's Dilemma]
\label{prop-physical-prisoners-dilemma}

Under the energy-based payoff model with physical constraints ($E_R > 0$, $c \ll E_R$, $\theta < (1/2 + c/E_R)/(1 + (1+\kappa)\delta^{\text{off}})$, $1 < \Phi < \Phi_{\text{PD}} = 2 + 4\theta(1+\kappa)\delta^{\text{def}}$), the single-round game has the structure of a Prisoner's Dilemma: $T > R > P > S$.

In the single-round game, defection is the dominant strategy for each agent. The unique Nash Equilibrium is $(D, D)$ with payoffs $(P, P)$, even though $(C, C)$ with payoffs $(R, R)$ is Pareto-superior.
\end{proposition}

\begin{proof}
The inequalities $T > R > P > S$ are verified above under the stated parameter conditions. In a simultaneous single-round game:

\begin{itemize}
\item If $B$ plays $C$: $A$ prefers $D$ ($T > R$).
\item If $B$ plays $D$: $A$ prefers $D$ ($P > S$).
\end{itemize}

So $D$ strictly dominates $C$ for each player. The unique NE is $(D, D)$. But $R > P$ means both agents would be strictly better off at $(C, C)$ than at $(D, D)$, establishing the Pareto inefficiency of the one-shot equilibrium.
\end{proof}

\textbf{Physical interpretation:} Each agent's individually rational
choice (defect regardless of the opponent's action) leads to a
collectively irrational outcome. The inefficiency is stronger than in
the standard abstract PD: because \(P < 0\) when \(\Phi > 2\) (the
physically realistic regime), mutual defection does not merely yield a
smaller gain but an absolute energy loss. Both agents would prefer the
cooperative outcome \((R, R)\), yet neither can unilaterally switch to
\(C\) without risking the sucker's payoff \(S < P < 0\). The one-shot
game therefore traps both agents in a net-negative equilibrium that the
repeated game (¬ß\ref{the-repeated-game-cooperation-as-equilibrium})
resolves.

\subsection{Numerical Illustration: The Baseline
Game}\label{numerical-illustration-the-baseline-game}

\textbf{Parameters:} \(E_R = 100\), \(c = 2\), \(\theta = 0.15\),
\(\delta^{\text{off}} = 0.15\), \(\delta^{\text{def}} = 0.25\),
\(\kappa = 2\).

Computed payoffs:

\[T = 100[1 - 0.15(1 + 3 \times 0.15)] = 100[1 - 0.15 \times 1.45] = 100 \times 0.7825 = 78.25\]

\[R = 50 - 2 = 48\]

\[\Phi = 1 + 3(0.15 + 0.25) = 2.2\]

\[P = 50(1 - 2.2/2) = 50(1 - 1.1) = 50 \times (-0.1) = -5\]

\[S = -0.15 \times 100 \times 3 \times 0.25 = -11.25\]

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}lll@{}}
\toprule\noalign{}
& \(B: C\) & \(B: D\) \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\(A: C\) & \textbf{(48, 48)} & (‚àí11.25, 78.25) \\
\(A: D\) & (78.25, ‚àí11.25) & (‚àí5, ‚àí5) \\
\end{longtable}
}

\textbf{Observations:}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Mutual defection is net-negative} (\(P = -5\)): both agents
  \emph{lose} energy. The conflict destroys more than the resource
  provides (consistent with Theorem 6 \citep{TC-IV}, since
  \(\Phi = 2.2 > 2\)).
\item
  \textbf{Mutual cooperation yields a large surplus} (\(R = 48\) each,
  system total = 96).
\item
  \textbf{Temptation is high} (\(T = 78.25\)), but the victim suffers
  (\(S = -11.25\)).
\item
  \textbf{The cooperation premium} is \(R - P = 48 - (-5) = 53\) per
  agent, the energy saved by cooperative behavior compared to the
  conflict equilibrium.
\end{enumerate}

\section{The Repeated Game: Cooperation as
Equilibrium}\label{the-repeated-game-cooperation-as-equilibrium}

\subsection{Why Repetition Matters}\label{why-repetition-matters}

The single-round Prisoner's Dilemma has defection as the unique NE. But
entities in the real world (biological organisms, economic agents,
nations) interact \textbf{repeatedly} over indefinite timeframes. The
theory of repeated games \citep{Fudenberg1986} shows that sustained
cooperation can emerge as an equilibrium in the repeated version of the
game.

In the standard (abstract-utility) PD, this result is qualitative:
cooperation \emph{can} be sustained for sufficiently patient agents, but
the critical patience threshold \(\delta^*\) depends on payoff values
that are free parameters. In our physical PD, the threshold is
\emph{structurally constrained}. Because mutual defection is
net-negative (\(P < 0\) when \(\Phi > 2\)), the punishment for defection
is automatically severe, and the resulting \(\delta^*\) is low. The
repeated-game analysis that follows is standard in its mechanism
(trigger strategies, folk theorem) but novel in its conclusion: the
physics sets the threshold, not the modeler.

The key parameter is the \textbf{discount factor} \(\delta \in (0, 1)\),
representing how much agents value future payoffs relative to present
ones. In our thermodynamic framework, \(\delta\) has a natural physical
interpretation:

\begin{definition}[Discount Factor]
\label{def-discount-factor}

The discount factor $\delta$ for an agent with planning horizon $T$ (periods) and per-period survival probability $\sigma \in (0, 1]$ is:

$$\delta = \sigma \cdot e^{-r}$$

where $r \geq 0$ is the time-preference rate. For entities optimizing over long horizons (high $\sigma$, low $r$), $\delta \to 1$. For entities with short horizons (low $\sigma$ or high $r$), $\delta \to 0$.
\end{definition}

\textbf{Physical interpretation:} \(\delta\) close to 0 means the agent
has a short planning horizon and discounts future payoffs heavily.
\(\delta\) close to 1 means the agent expects many future interactions
and weights them nearly equally with the present; this is formalized in
¬ß\ref{extension-to-n-player-games} below.

\subsection{Trigger Strategies: Grim and
Tit-for-Tat}\label{trigger-strategies-grim-and-tit-for-tat}

The simplest mechanism to sustain cooperation in repeated games is a
\textbf{trigger strategy}, a rule that starts cooperating and retaliates
against defection.

\begin{definition}[Grim Trigger]
\label{def-grim-trigger}

Agent $i$ plays $C$ in every period as long as no agent has ever played $D$. If any defection occurs, agent $i$ plays $D$ forever.
\end{definition}

\begin{definition}[Tit-for-Tat (TFT)]
\label{def-tit-for-tat}

Agent $i$ plays $C$ in period 1. In period $t > 1$, agent $i$ plays whatever the opponent played in period $t - 1$.
\end{definition}

Both strategies embody a simple enforcement mechanism: cooperation is
maintained as long as agents observe cooperation; defection triggers
retaliation. The threat of retaliation, sustained loss of cooperative
payoffs, creates a cost to defection that extends beyond the single
interaction.

\subsection{The Cooperation Condition (Grim
Trigger)}\label{the-cooperation-condition-grim-trigger}

Under the grim trigger strategy, the following holds:

\begin{theorem}[Cooperation as Nash Equilibrium: The Cooperation Theorem]
\label{thm-cooperation-theorem}

In the infinitely repeated energy-based game with discount factor $\delta$, the cooperative outcome $(C, C)$ in every period is sustainable as a Nash Equilibrium (under grim trigger) if and only if:

$$\delta \geq \frac{T - R}{T - P} = \delta^*$$

where $T$, $R$, $P$ are the temptation, reward, and punishment payoffs. The critical discount factor $\delta^*$ has the physical interpretation: cooperation is sustainable when agents sufficiently value future interactions.
\end{theorem}

\begin{proof}
Consider agent $A$'s incentive to deviate from the cooperative profile $(C, C, C, \ldots)$.

\textbf{Payoff from continued cooperation:}

$$V_A^{\text{coop}} = R + \delta R + \delta^2 R + \cdots = \frac{R}{1 - \delta}$$

\textbf{Payoff from one-shot defection (then punished forever):}

The deviating agent gets $T$ in the current period, then $(D, D)$ forever:

$$V_A^{\text{deviate}} = T + \delta P + \delta^2 P + \cdots = T + \frac{\delta P}{1 - \delta}$$

Cooperation is incentive-compatible when $V_A^{\text{coop}} \geq V_A^{\text{deviate}}$:

$$\frac{R}{1 - \delta} \geq T + \frac{\delta P}{1 - \delta}$$

$$R \geq (1 - \delta) T + \delta P$$

$$R - P \geq (1 - \delta)(T - P)$$

$$\frac{R - P}{T - P} \geq 1 - \delta$$

$$\delta \geq 1 - \frac{R - P}{T - P} = \frac{T - R}{T - P}$$

Since both agents face the same calculation (the game is symmetric), $(C, C)$ sustained by grim trigger is a NE whenever $\delta \geq \delta^*$.
\end{proof}

\subsection{\texorpdfstring{Computing \(\delta^*\) Under Physical
Constraints}{Computing \textbackslash delta\^{}* Under Physical Constraints}}\label{computing-delta-star}

Using the energy-based payoffs from
¬ß\ref{parameterization-the-physical-prisoners-dilemma}:

\[\delta^* = \frac{T - R}{T - P}\]

\textbf{Baseline numerical example}
(¬ß\ref{numerical-illustration-the-baseline-game} parameters):

\[\delta^* = \frac{78.25 - 48}{78.25 - (-5)} = \frac{30.25}{83.25} = 0.3634\]

\begin{corollary}[Low Cooperation Threshold]
\label{cor-low-cooperation-threshold}

Under the physical payoff model with $\Phi > 2$ (net-negative mutual defection), the critical discount factor satisfies $\delta^* < 0.5$. Agents need only value the future at more than 36% of the present to sustain cooperation.
\end{corollary}

\begin{proof}
When $P < 0$ (which occurs for $\Phi > 2$, per the baseline illustration, ¬ß\ref{numerical-illustration-the-baseline-game}):

$$T - P = T + |P| > T$$

$$\delta^* = \frac{T - R}{T - P} < \frac{T - R}{T} = 1 - \frac{R}{T}$$

It remains to show $R > T/2$, i.e., $1 - R/T < 0.5$. Substituting the payoff expressions from the physical parameterization (¬ß\ref{parameterization-the-physical-prisoners-dilemma}), $R = E_R/2 - c$ and $T = E_R[1 - \theta(1 + (1+\kappa)\delta^{\text{off}})]$, the condition $R > T/2$ reduces to:

$$c < \frac{\theta E_R}{2}\bigl(1 + (1+\kappa)\delta^{\text{off}}\bigr)$$

This holds whenever the exploitation overhead exceeds twice the coordination cost, a condition satisfied for any non-trivial $\theta$, since the PD ordering already requires $c \ll E_R$. For the baseline values ($c = 2$, $\theta = 0.15$, $E_R = 100$, $\kappa = 2$, $\delta^{\text{off}} = 0.15$): the right side is $0.15 \times 100 \times 1.45 / 2 = 10.875 \gg 2 = c$, and $R/T = 48/78.25 = 0.614 > 0.5$, giving $\delta^* < 1 - 0.614 = 0.386 < 0.5$.
\end{proof}

\textbf{Physical interpretation:} Because mutual defection under
physical energy payoffs is net-negative (\(P < 0\)), the punishment for
defection is \textbf{automatically severe}: agents do not just earn
less, they actively \emph{lose} energy. This makes cooperation easy to
sustain: even agents with modest foresight (\(\delta > 0.36\)) find it
in their interest to cooperate. Compare this with the abstract PD (e.g.,
\(T=5, R=3, P=1, S=0\)) where \(\delta^* = 0.5\). Physical constraints
make cooperation \textbf{more} accessible, not less.

\subsection{\texorpdfstring{Sensitivity Analysis: \(\delta^*\) Across
Parameter
Regimes}{Sensitivity Analysis: \textbackslash delta\^{}* Across Parameter Regimes}}\label{sensitivity-analysis-delta-star}

The friction-regime rows below vary the repair multiplier \(\kappa\)
(the physical knob from the Second Law: see Theorem 6 \citep{TC-IV})
holding \(\delta^{\text{off}} = 0.15\), \(\delta^{\text{def}} = 0.25\),
\(\theta = 0.15\), \(c = 2\), \(E_R = 100\). Because \(\Phi\), \(T\),
and \(S\) all share the factor \((1+\kappa)\), they are recomputed
consistently for each row. The exploitation-cost rows hold
\(\kappa = 2\) and vary \(\theta\).

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.2609}}
  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.0435}}
  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.0870}}
  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.0870}}
  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.0870}}
  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.0870}}
  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.0870}}
  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.1304}}
  >{\raggedright\arraybackslash}p{(\linewidth - 16\tabcolsep) * \real{0.1304}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Regime
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(\kappa\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(\theta\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(\Phi\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(T\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(R\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(P\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(S\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(\delta^*\)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
No repair cost & 0 & 0.15 & 1.4 & 82.75 & 48 & 15 & ‚àí3.75 & 0.513 \\
Low repair & 1 & 0.15 & 1.8 & 80.50 & 48 & 5 & ‚àí7.50 & 0.430 \\
\textbf{Baseline} & 2 & 0.15 & 2.2 & 78.25 & 48 & ‚àí5 & ‚àí11.25 &
\textbf{0.363} \\
High repair & 4 & 0.15 & 3.0 & 73.75 & 48 & ‚àí25 & ‚àí18.75 & 0.261 \\
Very high & 6 & 0.15 & 3.8 & 69.25 & 48 & ‚àí45 & ‚àí26.25 & 0.186 \\
Cheap exploit & 2 & 0.05 & 2.2 & 92.75 & 48 & ‚àí5 & ‚àí3.75 & 0.458 \\
Expensive exploit & 2 & 0.30 & 2.2 & 56.50 & 48 & ‚àí5 & ‚àí22.50 & 0.138 \\
\end{longtable}
}

\begin{quote}
\textbf{Note on game type.} The PD/Chicken boundary is
\(\Phi_{\text{PD}} = 2 + 4\theta(1+\kappa)\delta^{\text{def}}\). The
high-repair rows (\(\kappa = 4\) with
\(\Phi = 3.0 > \Phi_{\text{PD}} = 2.75\); \(\kappa = 6\) with
\(\Phi = 3.8 > \Phi_{\text{PD}} = 3.05\)) and the cheap-exploit row
(\(\Phi_{\text{PD}} = 2 + 4(0.05)(3)(0.25) = 2.15 < 2.2\)) are Chicken
games (\(T > R > S > P\)), not Prisoner's Dilemmas. The cooperation
threshold formula \(\delta^* = (T - R)/(T - P)\) and all subsequent
theorems apply identically to both game structures; the Chicken
structure actually yields a \emph{lower} \(\delta^*\) because \(T - P\)
is larger.
\end{quote}

\textbf{Key findings:}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Higher friction ‚Üí lower \(\delta^*\) ‚Üí easier cooperation.}
  When conflict is more destructive, the ``punishment'' for mutual
  defection is harsher, making cooperation sustainable even for
  short-sighted agents.
\item
  \textbf{Higher exploitation cost ‚Üí lower \(\delta^*\).} When defection
  is expensive (well-defended targets), the temptation premium shrinks
  and cooperation is easier.
\item
  \textbf{In all physical regimes, \(\delta^* < 0.6\).} Cooperation is
  sustainable for any agent that values future interactions at more than
  \textasciitilde60\% of present ones, a modest requirement.
\end{enumerate}

\section{Network Effects: The Friction
Amplifier}\label{network-effects-the-friction-amplifier}

\subsection{Incorporating Network Friction into
Payoffs}\label{incorporating-network-friction-into-payoffs}

The single-interaction payoffs above omit a crucial physical cost:
defection injects friction into the network (Definition 8
\citep{TC-IV}), degrading all future cooperative transactions for every
agent in the network. This makes defection costlier than the
single-round analysis suggests, and the additional cost enters the
repeated game through the punishment branch of the equilibrium condition
rather than the single-round payoff matrix.

\subsection{The Full Defection Cost}\label{the-full-defection-cost}

Over the infinite horizon, the total cost of a single defection of
severity \(v\) includes the cascading friction (Theorem 10
\citep{TC-IV}):

\[C_{\text{cascade}} = \frac{\eta \cdot v \cdot M \cdot \bar{E}}{\epsilon}\]

This cost is distributed across all \(M\) transactions and all agents in
the network. The fraction borne by the defecting agent (through degraded
future cooperative payoffs) is:

\[C_{\text{defector}}^{\text{cascade}} = \frac{C_{\text{cascade}}}{N} = \frac{\eta \cdot v \cdot M \cdot \bar{E}}{N \cdot \epsilon}\]

\begin{theorem}[Network-Adjusted Cooperation Condition]
\label{thm-network-adjusted-cooperation}

When network friction effects are included, the effective punishment payoff becomes:

$$\tilde{P} = P - \frac{C_{\text{defector}}^{\text{cascade}}}{(1/\epsilon)} = P - \frac{\eta \cdot v \cdot M \cdot \bar{E}}{N}$$

(amortized over the recovery time horizon). The network-adjusted critical discount factor is:

$$\tilde{\delta}^* = \frac{T - R}{T - \tilde{P}} < \delta^*$$

Network friction makes cooperation strictly easier to sustain.
\end{theorem}

\begin{proof}
Since the cascading friction is a real energy cost borne by the defector (via degraded future cooperative opportunities), and $C_{\text{defector}}^{\text{cascade}} > 0$, we have $\tilde{P} < P$. Therefore $T - \tilde{P} > T - P$, and:

$$\tilde{\delta}^* = \frac{T - R}{T - \tilde{P}} < \frac{T - R}{T - P} = \delta^*$$
\end{proof}

\subsection{Numerical Illustration}\label{numerical-illustration}

For the violation severity we set \(v = E_R\), the per-interaction
resource pool. This represents a full-violation baseline: a defector who
appropriates the counterpart's entire resource share causes harm equal
to \(E_R\). This is the natural unit for energy-denominated violations,
and the bound \(v \leq E_R\) follows from resource conservation (a
single interaction cannot extract more than \(E_R\) from the system).
The numerical results scale linearly with \(v\); smaller violations
reduce \(C_{\text{cascade}}\) proportionally.

Using baseline parameters (\(E_R = 100\), \(M = 1000\),
\(\bar{E} = 10\), \(\eta = 0.01\), \(v = E_R = 100\), \(N = 100\),
\(\epsilon = 0.05\)):

\[C_{\text{defector}}^{\text{cascade}} = \frac{0.01 \times 100 \times 1000 \times 10}{100 \times 0.05} = \frac{10{,}000}{5} = 2{,}000\]

Amortized over the recovery period (\(\sim 1/\epsilon = 20\) periods):

\[\tilde{P}_{\text{per period}} \approx P - 2000/20 = -5 - 100 = -105\]

\[\tilde{\delta}^* = \frac{78.25 - 48}{78.25 - (-105)} = \frac{30.25}{183.25} = 0.165\]

\textbf{With network friction, the cooperation threshold drops to
\(\tilde{\delta}^* = 0.165\).} Even extremely short-sighted agents find
cooperation rational when the network punishes defection through
friction cascades.

\subsection{Deception Penalties}\label{deception-penalties}

Defection via strategic signal corruption rather than overt conflict
incurs the micro-friction costs derived in \citep{TC-II}:

\begin{itemize}
\tightlist
\item
  Decision cost to the victim: \(\Delta U_B(q)\) (Theorem 1
  \citep{TC-II})
\item
  Verification overhead on the network: \(\Delta W(q)\) (Theorem 2
  \citep{TC-II})
\item
  Deception-induced friction:
  \(\Delta\phi_{\text{info}} = \eta_{\text{info}} \cdot \Delta H \cdot W_{\text{bit}}\)
  (Definition 6 \citep{TC-II})
\end{itemize}

These costs further increase the effective punishment and reduce
\(\tilde{\delta}^*\). The deception channel does not create a
``cheaper'' form of defection; it incurs different but equally real
thermodynamic costs.

\section{\texorpdfstring{Extension to \(N\)-Player
Games}{Extension to N-Player Games}}\label{extension-to-n-player-games}

\subsection{\texorpdfstring{The \(N\)-Player Public Goods
Game}{The N-Player Public Goods Game}}\label{the-n-player-public-goods-game}

To extend beyond two players, consider the canonical \(N\)-player
interaction: a \textbf{public goods game} with energy-denominated
payoffs.

\begin{definition}[$N$-Player Energy Public Goods Game]
\label{def-n-player-public-goods}

$N$ agents each choose whether to Cooperate ($C$: contribute energy $c_i > 0$ to a shared pool) or Defect ($D$: contribute nothing). The shared pool generates a return multiplied by factor $\alpha > 1$ (the efficiency gain from cooperation) and is distributed equally. Agent $i$'s payoff:

$$\pi_i = \frac{\alpha \sum_{j=1}^{N} c_j \cdot \mathbb{1}[j \text{ cooperates}]}{N} - c_i \cdot \mathbb{1}[i \text{ cooperates}] - \phi \cdot \bar{E} \cdot \frac{N - n_C}{N}$$

where $n_C$ is the number of cooperators and the last term captures the friction tax from defectors (proportional to the defection rate).
\end{definition}

In our framework, \(\alpha\) is the \textbf{cooperative multiplier}, the
ratio of total output under cooperation to total input. This is derived
from the Lagrangian constraints result that the variational equilibrium
allocates resources more efficiently than the unconstrained regime.

More concretely, when all \(N\) agents cooperate, each contributes \(c\)
and receives \(\alpha c\) (net gain \((\alpha - 1)c\)). When \(n_C < N\)
agents cooperate, the cooperators pay the coordination cost while
defectors free-ride, but the system suffers friction loss.

\subsection{\texorpdfstring{The One-Shot \(N\)-Player
Equilibrium}{The One-Shot N-Player Equilibrium}}\label{the-one-shot-n-player-equilibrium}

In the single-round game, defection is dominant when:

\[\frac{\alpha c}{N} < c \iff \alpha < N\]

That is, defection dominates when the cooperative return to each
individual is less than the individual's contribution, the standard
free-rider problem. For large \(N\), this holds unless \(\alpha\) is
very large.

The one-shot NE is universal defection, yielding payoff
\(\pi_i^{DD} = -\phi_0 \bar{E}\) (only the friction tax from the
non-cooperative environment). The cooperative payoff is
\(\pi_i^{CC} = (\alpha - 1)c\). The efficiency loss is:

\[\Delta\pi = (\alpha - 1)c + \phi_0 \bar{E} > 0\]

\subsection{\texorpdfstring{The Repeated \(N\)-Player
Game}{The Repeated N-Player Game}}\label{the-repeated-n-player-game}

\begin{theorem}[N-Player Cooperation Sustainability]
\label{thm-n-player-cooperation}

In the infinitely repeated $N$-player energy public goods game with discount factor $\delta$ and grim trigger punishment (all agents defect forever upon any defection), cooperation is a NE if:

$$\delta \geq \delta_N^* = \frac{\pi_i^{\text{deviate}} - \pi_i^{CC}}{\pi_i^{\text{deviate}} - \pi_i^{DD}}$$

where $\pi_i^{\text{deviate}}$ is agent $i$'s payoff from unilaterally defecting while all others cooperate:

$$\pi_i^{\text{deviate}} = \frac{\alpha (N-1) c}{N} - 0 = \frac{\alpha(N-1)c}{N}$$

(receives the public good without contributing). The critical discount factor is:

$$\delta_N^* = \frac{\frac{\alpha(N-1)c}{N} - (\alpha - 1)c}{\frac{\alpha(N-1)c}{N} + \phi_0 \bar{E}} = \frac{c(N - \alpha)}{(N-1)\alpha c + N\phi_0\bar{E}}$$
\end{theorem}

\begin{proof}
The deviation payoff is $\pi_i^{\text{deviate}} = \alpha(N-1)c/N$ (receives share of the $(N-1)$ cooperators' contributions without paying). The cooperative payoff is $\pi_i^{CC} = \alpha c - c = (\alpha-1)c$. The punishment payoff is $\pi_i^{DD} = -\phi_0\bar{E}$ (universal defection with friction tax). Applying the grim trigger calculation:

$$\delta_N^* = \frac{\pi_i^{\text{deviate}} - \pi_i^{CC}}{\pi_i^{\text{deviate}} - \pi_i^{DD}} = \frac{\alpha(N-1)c/N - (\alpha-1)c}{\alpha(N-1)c/N + \phi_0\bar{E}}$$

Simplifying the numerator:

$$\frac{\alpha(N-1)c - N(\alpha-1)c}{N} = \frac{c[\alpha N - \alpha - \alpha N + N]}{N} = \frac{c(N - \alpha)}{N}$$

So:

$$\delta_N^* = \frac{c(N - \alpha)/N}{\alpha(N-1)c/N + \phi_0\bar{E}} = \frac{c(N-\alpha)}{\alpha(N-1)c + N\phi_0\bar{E}}$$
\end{proof}

\emph{Remark:} The deviation payoff above omits the friction tax
\(\phi_0\bar{E}/N\) that Definition~\ref{def-n-player-public-goods}
imposes on all agents, including the deviator, in the defection round.
Including this term yields the corrected threshold
\(\delta_N^{*,\text{corr}} = [c(N-\alpha) - \phi_0\bar{E}] / [(N-1)(\alpha c + \phi_0\bar{E})]\),
which is strictly smaller than the stated \(\delta_N^*\). The omission
is therefore conservative: cooperation is achievable for any
\(\delta \geq \delta_N^*\) per the theorem, and in fact for the strictly
smaller \(\delta_N^{*,\text{corr}}\) as well. All qualitative
conclusions (cooperation scales with \(N\), friction lowers the
threshold) are strengthened rather than weakened.

\subsection{\texorpdfstring{Properties of
\(\delta_N^*\)}{Properties of \textbackslash delta\_N\^{}*}}\label{properties-of-delta-n-star}

\begin{corollary}[Cooperation Strengthens with Network Friction]
\label{cor-cooperation-strengthens-friction}

$$\frac{\partial \delta_N^*}{\partial \phi_0} < 0$$

Higher network friction (harsher punishment for the all-defect state) makes cooperation easier to sustain.
\end{corollary}

\begin{proof}
The numerator $c(N - \alpha)$ does not depend on $\phi_0$. The denominator $\alpha(N-1)c + N\phi_0\bar{E}$ is strictly increasing in $\phi_0$. Therefore $\delta_N^*$ is strictly decreasing in $\phi_0$.
\end{proof}

\begin{corollary}[Cooperation Strengthens with Scale Under Friction]
\label{cor-cooperation-strengthens-scale}

When friction is non-negligible ($\phi_0 > 0$), the critical discount factor $\delta_N^*$ remains bounded (does not approach 1) as $N \to \infty$:

$$\lim_{N \to \infty} \delta_N^* = \frac{c}{\alpha c + \phi_0 \bar{E}} < 1$$
\end{corollary}

\begin{proof}
Dividing numerator and denominator by $N$:

$$\delta_N^* = \frac{c(1 - \alpha/N)}{\alpha(1 - 1/N)c + \phi_0\bar{E}} \xrightarrow{N \to \infty} \frac{c}{\alpha c + \phi_0\bar{E}}$$

Since $\phi_0 > 0$: $\alpha c + \phi_0\bar{E} > c$ (for $\alpha > 1$ or $\phi_0 > 0$), so the limit is strictly less than 1.
\end{proof}

\textbf{Physical interpretation:} In classical game theory (abstract
payoffs, no friction), the critical discount factor for \(N\)-player
public goods games approaches 1 as \(N \to \infty\), so cooperation
becomes unsustainable in large groups. With energy-based payoffs and
network friction, this pathology \textbf{disappears}: the friction
penalty of universal defection grows with group size, keeping
cooperation accessible even in large populations. This is the formal
proof that the cooperative equilibrium holds and strengthens as the
network scales.

\subsection{\texorpdfstring{Numerical Illustration: \(N\)-Player
Cooperation
Thresholds}{Numerical Illustration: N-Player Cooperation Thresholds}}\label{n-player-cooperation-thresholds}

\textbf{Parameters:} \(c = 10\), \(\alpha = 2\) (cooperative doubling),
\(\phi_0 = 0.5\) (moderate background friction), \(\bar{E} = 10\).

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\linewidth - 6\tabcolsep) * \real{0.2500}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\(N\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Deviation gain \(\pi^{\text{dev}} - \pi^{CC}\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Punishment gap \(\pi^{\text{dev}} - \pi^{DD}\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(\delta_N^*\)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
2 & 0 & 10 + 5 = 15 & 0.000 \\
3 & 3.33 & 13.33 + 5 = 18.33 & 0.182 \\
5 & 6.00 & 16 + 5 = 21 & 0.286 \\
10 & 8.00 & 18 + 5 = 23 & 0.348 \\
50 & 9.60 & 19.6 + 5 = 24.6 & 0.390 \\
100 & 9.80 & 19.8 + 5 = 24.8 & 0.395 \\
\(\infty\) & 10.00 & 20 + 5 = 25 & 0.400 \\
\end{longtable}
}

\textbf{Key result:} Even as \(N \to \infty\), \(\delta_N^*\) converges
to 0.4, well below 1. Cooperation remains a Nash Equilibrium for any
agent with \(\delta \geq 0.4\). Without friction (\(\phi_0 = 0\)),
\(\delta_N^*\) for \(N = 100\) would be \(9.8/19.8 = 0.495\), and for
\(N \to \infty\) it would approach \(10/20 = 0.5\). Friction provides
the extra margin that makes cooperation robust.

\section{Evolutionary Stability: Defection Cannot
Invade}\label{evolutionary-stability-defection-cannot-invade}

\subsection{The Evolutionary Game-Theoretic
Framework}\label{the-evolutionary-game-theoretic-framework}

Beyond asking whether cooperation is a NE, we ask: \textbf{can defection
invade a cooperative population?} This is the evolutionary stability
question \citep{MaynardSmith1973}.

\begin{definition}[Evolutionary Stable Strategy (ESS)]
\label{def-ess}

A strategy $s^*$ is an ESS if, for any mutant strategy $s' \neq s^*$:

(a) $\pi(s^*, s^*) > \pi(s', s^*)$ (\textbf{strict NE}), or

(b) $\pi(s^*, s^*) = \pi(s', s^*)$ and $\pi(s^*, s') > \pi(s', s')$ (\textbf{stability condition}).
\end{definition}

\subsection{TFT as an Approximate ESS}\label{tft-as-an-approximate-ess}

In the iterated Prisoner's Dilemma with energy payoffs, Tit-for-Tat
(TFT) is a well-known strong performer \citep{Axelrod1984}. While TFT is
not a strict ESS in the technical sense (it does not strictly outperform
other ``nice'' strategies like Always-Cooperate in a homogeneous
population), it satisfies a strong \textbf{invasion barrier} condition:

\begin{theorem}[Defection Invasion Barrier]
\label{thm-defection-invasion-barrier}

In a population of $N$ TFT-playing agents with discount factor $\delta > \delta^*$, a mutant Always-Defect (ALLD) agent earns:

$$V_{\text{ALLD}} = T + \frac{\delta P}{1 - \delta}$$

while a TFT agent in the same population earns:

$$V_{\text{TFT}} = \frac{R}{1 - \delta}$$

The ALLD mutant is strictly dominated ($V_{\text{ALLD}} < V_{\text{TFT}}$) when:

$$\delta > \frac{T - R}{T - P} = \delta^*$$

Under the physical payoff model with $\Phi > 2$, this holds for all $\delta > 0.363$ (baseline parameters).
\end{theorem}

\begin{proof}
An ALLD mutant, upon meeting a TFT agent, defects in round 1 (getting $T$) and then faces retaliation: mutual defection from round 2 onward (getting $P$ each round). Its lifetime payoff against any TFT agent is:

$$V_{\text{ALLD}} = T + \delta P + \delta^2 P + \cdots = T + \frac{\delta P}{1 - \delta}$$

A TFT agent meeting other TFT agents cooperates every round:

$$V_{\text{TFT}} = R + \delta R + \delta^2 R + \cdots = \frac{R}{1 - \delta}$$

Then $V_{\text{TFT}} > V_{\text{ALLD}}$:

$$\frac{R}{1 - \delta} > T + \frac{\delta P}{1 - \delta}$$

$$R > (1 - \delta)T + \delta P$$

$$\delta > \frac{T - R}{T - P} = \delta^*$$

which is exactly the cooperation condition from Theorem~\ref{thm-cooperation-theorem}.
\end{proof}

\textbf{Physical interpretation:} A defector entering a cooperative
population gets one ``hit'' (the temptation payoff \(T\)) and then finds
itself in perpetual mutual defection, which under physical payoffs means
perpetual net-negative returns (\(P < 0\)). The defector's strategy is
self-defeating: it destroys the cooperative environment it needs to
profit from. In evolutionary terms, the ``invasion fitness'' of
defection is negative in a cooperative population.

\subsection{The Defector's Dilemma: Extinction Through
Friction}\label{the-defectors-dilemma-extinction-through-friction}

Beyond the direct retaliation mechanism, the defector also faces the
cascading friction cost (¬ß\ref{network-effects-the-friction-amplifier}).
Even if a defector can avoid direct retaliation (e.g., in a large
anonymous population), each defection increases the network friction
\(\phi\), which degrades the returns to \emph{all} agents, including the
defector. In a population where the defector's payoff depends on the
cooperative surplus (because even defectors benefit from a low-friction
network for their non-defection interactions), systematic defection is a
self-undermining strategy.

\begin{corollary}[Defection Self-Extinction in Friction Environments]
\label{cor-defection-self-extinction}

In a network with friction coupling, a defector whose defection rate $\nu_D$ injects enough friction to push $\phi$ above the network collapse threshold (Corollary 11 \citep{TC-IV}) destroys the cooperative surplus from which it profits. In the limit, a population of defectors achieves the minimum network payoff $\pi^{DD} = -\phi_0 \bar{E}$, which is strictly less than zero. Defection is not merely dominated; it is self-destructive.
\end{corollary}

\section{Tit-for-Tat Dynamics in Energy
Terms}\label{tit-for-tat-dynamics-in-energy-terms}

\subsection{Short-Term Defection Profit is
Erasure}\label{short-term-defection-profit-is-erasure}

We now formally show that a defector's short-term energy gain is
\textbf{wiped out} over repeated interactions.

\begin{theorem}[Defection Profit Erasure]
\label{thm-defection-profit-erasure}

An agent that defects once against a TFT population and then returns to cooperation incurs a net lifetime loss whenever:

$$\delta > \frac{T - R}{R - S}$$

(without friction). Under the physical payoff model (baseline parameters), the basic threshold is $\delta = 0.511$. When network friction costs are included ($F = 100$ per period over 20 recovery periods), the threshold drops to $\delta \approx 0.17$.
\end{theorem}

\begin{proof}
Consider an agent that defects in period $t$ (getting $T$), is retaliated against in period $t+1$ (getting $S$), and then both agents return to cooperation from period $t+2$ onward. The deviation costs the agent:

\begin{itemize}
\item Period $t$: gains $T$ instead of $R$ (net gain: $T - R$)
\item Period $t+1$: earns $S$ instead of $R$ (net loss: $R - S$)
\item Periods $t+2, \ldots$: back to $R$ (no change)
\end{itemize}
Present-value net gain from the one-shot deviation:

$$\Delta V = (T - R) - \delta(R - S)$$

The deviation is unprofitable ($\Delta V < 0$) when:

$$\delta > \frac{T - R}{R - S}$$

With baseline parameters: $\delta > (78.25 - 48)/(48 - (-11.25)) = 30.25/59.25 = 0.511$.

However, this analysis ignores the network friction cost. Including the per-period friction penalty amortized over recovery time ($\sim 1/\epsilon = 20$ periods), even a single defection imposes a sustained future cost. From ¬ß\ref{numerical-illustration}, the per-defector cascade cost is $C_{\text{defector}}^{\text{cascade}} = 2{,}000$, giving $F = 2{,}000 / 20 = 100$ energy units per period of degraded cooperative returns. When we include this:

$$\Delta V_{\text{adj}} = (T - R) - \delta(R - S) - \delta \sum_{k=1}^{1/\epsilon} \delta^{k-1} F$$

This makes the defection-erasure threshold strictly lower than $\delta = 0.511$:

$$\frac{T - R}{R - S + \delta F/(1-\delta)} < \frac{T - R}{R - S}$$

Numerically, the condition $\Delta V_{\text{adj}} < 0$ holds for $\delta \gtrsim 0.17$ after including friction.
\end{proof}

\textbf{Physical interpretation:} The defector's profit
(\(T - R = 30.25\)) is a one-time energy windfall. The retaliation cost
(\(R - S = 59.25\)) comes one period later. The friction cost
(\(F = 100\)/period) compounds over \(\sim 20\) periods. For any agent
that expects to survive more than a few interactions
(\(\delta > 0.17\)), the short-term defection profit is negative on net.
Defection is an energetic ``payday loan'' with punishing interest.

\section{System-Level Optimality: The Welfare
Theorem}\label{system-level-optimality-the-welfare-theorem}

\subsection{The Social Welfare
Function}\label{the-social-welfare-function}

Define the \textbf{system welfare} as the total net energy available to
all agents after interaction costs:

\[SW = \sum_{i=1}^{N} \pi_i - \mathcal{C}_{\text{total}}\]

where \(\mathcal{C}_{\text{total}}\) is the total interaction cost
(Definition 11 \citep{TC-IV} plus Definition 7 \citep{TC-II}).

\subsection{The Maximum Welfare
Theorem}\label{the-maximum-welfare-theorem}

\begin{theorem}[Cooperation Maximizes System Welfare]
\label{thm-cooperation-maximizes-welfare}

Among all strategy profiles in the repeated $N$-player game, the cooperative profile ($C$ for all agents in all periods) uniquely maximizes the per-period system welfare:

$$SW^{CC} = N(\alpha - 1)c \geq SW^{s} \quad \forall \text{ strategy profiles } s$$

with equality only when $s$ is also universally cooperative.
\end{theorem}

\begin{proof}
Under universal cooperation:
\begin{itemize}
\item Total resource input: $Nc$ (all agents contribute)
\item Total output: $\alpha N c$ (cooperative multiplier)
\item Net welfare: $N(\alpha - 1)c$
\item Interaction cost: $\mathcal{C}_{\text{total}} = 0$ (no conflict, no deception, $\phi = 0$)
\end{itemize}
Under any profile with $n_D > 0$ defectors:
\begin{itemize}
\item Resource input: $(N - n_D)c < Nc$ (defectors do not contribute)
\item Output: $\alpha(N - n_D)c < \alpha Nc$ (reduced by lost contributions)
\item Network friction tax: $\phi \cdot M \cdot \bar{E} > 0$ (deviators bear friction penalty)
\item Net welfare: strictly less than $N(\alpha - 1)c$
\end{itemize}
Formally:

$$SW^{CC} - SW^{s} = \underbrace{n_D (\alpha - 1)c}_{\text{lost contributions}} + \underbrace{\phi(s) M \bar{E}}_{\text{friction tax}} > 0$$

for any $s$ with $n_D > 0$. Uniqueness follows from the strict inequality above: for any strategy profile with $n_D \geq 1$ defectors, the lost-contributions term $n_D(\alpha - 1)c > 0$ ensures that no defecting profile achieves $SW^{CC}$. (*Note: The Tullock contest dissipation term $D(s)$ applies to arms-race conflict settings (see Proposition 1 \citep{TC-IV}) and does not apply in this public goods free-rider context; uniqueness here rests solely on the lost-contributions and friction-tax terms.*)
\end{proof}

\subsection{The System-Level Nash
Equilibrium}\label{the-system-level-nash-equilibrium}

Combining Theorem~\ref{thm-cooperation-theorem} and
Theorem~\ref{thm-cooperation-maximizes-welfare}:

\begin{corollary}[Cooperation as the Unique Efficient Nash Equilibrium]
\label{cor-cooperation-unique-efficient-ne}

In the infinitely repeated energy-based game with $\delta > \delta^*$:

(a) Universal cooperation is a Nash Equilibrium (Theorem~\ref{thm-cooperation-theorem}).

(b) Universal cooperation uniquely maximizes system welfare (Theorem~\ref{thm-cooperation-maximizes-welfare}).

(c) Universal cooperation minimizes total interaction cost ($\mathcal{C}_{\text{total}} = 0$).

Therefore, the cooperative strategy profile ("cooperation") is the unique Nash Equilibrium that is also Pareto-optimal and system-welfare-maximizing.
\end{corollary}

\textbf{Physical interpretation:} Among all Nash Equilibria of the
repeated energy-denominated game, universal cooperation is the unique
profile that maximizes aggregate welfare and minimizes total
dissipation. The selection is not hypothetical; it is the standard Nash
equilibrium selection criterion applied to energy-denominated payoffs
under physical constraints.

\section{Discussion}\label{discussion}

\subsection{Physical Interpretation}\label{physical-interpretation}

The payoff matrix derived in ¬ß\ref{the-energy-denominated-payoff-model}
is determined entirely by the resource value \(E_R\), the exploitation
efficiency \(\theta\), the friction multiplier \(\Phi\), and the
coordination cost \(c\). The Prisoner's Dilemma ordering
\(T > R > P > S\) (Proposition~\ref{prop-physical-prisoners-dilemma})
holds in the parameter region where exploitation is cheap and friction
is moderate; at higher friction (\(\Phi \geq \Phi_{\text{PD}}\)) the
ordering becomes that of Hawk-Dove, with \(S > P\). Both orderings
emerge from the same physical model under different parameter values,
and the location of the transition is fixed by \citep{TC-IV} rather than
chosen.

The negativity of \(P\) under \(\Phi > 2\) is a direct consequence of
the Second Law as it acts on contested boundaries. Boundary structures
are organized configurations that required directed work to assemble; an
attack disrupts that organization in a single irreversible step, while
restoration requires sustained work against the entropic gradient. The
condition \(\Phi > 2\) states that the work cost of repair, amplified by
the irreversibility factor \(\kappa > 1\), exceeds the work value
extracted from the resource. Mutual defection is therefore
net-dissipative: more energy is converted to waste heat than is captured
as usable resource.

The cooperation threshold \(\delta^*\) admits a corresponding physical
reading. The discount factor \(\delta = \sigma e^{-r}\) encodes
per-period survival probability \(\sigma\) and time-preference rate
\(r\). An agent satisfying \(\delta \geq \delta^*\) is one whose
expected continuation in the system is sufficient that future energy
access dominates the immediate energy windfall of seizure. The threshold
\(\delta^* \approx 0.363\) is the level of expected persistence at which
an agent's own thermodynamic future outweighs the one-shot temptation;
the cooperation theorem (Theorem~\ref{thm-cooperation-theorem}) is the
statement that an agent intending to persist will not defect on a
cooperator.

The friction state variable \(\phi_t\) tracks accumulated entropy in
shared infrastructure. Each defection injects \(\Delta\phi = \eta v\)
into the network, and recovery proceeds at rate \(\epsilon\) per period
\citep{TC-IV}. The repeated-game payoff coupling through \(\phi_t\)
(Theorem~\ref{thm-network-adjusted-cooperation}) is the second-law cost
of restoration distributed across all participants: a defector exports
entropy onto the network, and subsequent cooperative transactions pay
the work cost of importing free energy to drive \(\phi_t\) back toward
baseline. The network-adjusted threshold
\(\tilde{\delta}^* \approx 0.165\) reflects this additional cost;
defection becomes more difficult to sustain because each instance
damages the substrate of all future interactions.

Welfare uniqueness (Theorem~\ref{thm-cooperation-maximizes-welfare},
Corollary~\ref{cor-cooperation-unique-efficient-ne}) follows from the
same dissipation accounting. Universal cooperation is the only profile
in which contest investment \(e^*\), boundary damage, and friction
injection are all zero. Every other profile expends energy on contest,
repair, or network restoration. The welfare-maximizing equilibrium is
therefore the profile that minimizes irreversible work per period, and
the result is not a coincidence of utility specification but a property
of the underlying energy accounting.

\subsection{Testable Predictions}\label{testable-predictions}

The model generates numerically specific, falsifiable predictions:

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}
  >{\raggedright\arraybackslash}p{(\linewidth - 4\tabcolsep) * \real{0.3333}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Prediction
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Formal source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Falsification criterion
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Cooperation threshold \(\delta^* \approx 0.36\) under baseline friction
& Theorem~\ref{thm-cooperation-theorem} & Cooperation thresholds in
energy-denominated experimental games consistently exceed \(0.5\) \\
Single-deviation profit is net-negative for \(\delta > 0.17\) &
Theorem~\ref{thm-defection-profit-erasure} & Single defectors in
iterated games with endogenous reputation costs show net-positive
lifetime payoffs \\
\(N\)-player cooperation threshold converges to \(\sim 0.4\) as
\(N \to \infty\) & Theorem~\ref{thm-n-player-cooperation} & Large-group
experiments with friction show \(\delta_N^* \to 1\) as in the standard
public goods game \\
Network collapse at critical violation frequency &
Corollary~\ref{cor-defection-self-extinction}, Corollary 11
\citep{TC-IV} & Institutional-trust data shows no threshold effect
relating violation frequency to cooperative breakdown \\
Defection invasion fitness is negative in cooperative populations &
Theorem~\ref{thm-defection-invasion-barrier} & Defector mutants
consistently invade and persist in cooperative populations under
physical payoff conditions \\
\end{longtable}
}

\subsection{Limitations}\label{limitations}

The game-theoretic core inherits standard simplifications from the
repeated-games literature:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Binary strategy space.} Agents choose \(C\) or \(D\). Real
  interactions admit continuous gradations of compliance. A continuous
  extension, where agents choose a degree of compliance
  \(\lambda \in [0,1]\) (distinct from the cooperative multiplier
  \(\alpha\) used in the \(N\)-player public goods game), would refine
  the payoff landscape but would not eliminate the friction asymmetry
  driving the result. We conjecture that the cooperative equilibrium
  persists and that Corollary~\ref{cor-cooperation-unique-efficient-ne}
  generalizes to a cooperation manifold.
\item
  \textbf{Indefinite horizon.} The cooperation results assume no
  anticipated final period. For agents with long planning horizons, this
  is well-matched; applications to agents with known finite lifespans
  require the finite-horizon extension with reputation effects.
\item
  \textbf{Perfect monitoring.} The grim trigger strategy assumes
  deviations are observed immediately. Under imperfect monitoring, the
  standard result is that cooperation remains a NE but requires a higher
  \(\delta > \delta^*_{\text{noisy}} > \delta^*\)
  \citep{GreenPorter1984, Fudenberg1991}. The model's baseline
  \(\delta^* = 0.363\) provides substantial headroom before cooperation
  ceases to be sustainable.
\end{enumerate}

\textbf{Scope.} The payoff model assumes symmetric agents contesting a
single resource per interaction. Asymmetric agents (different \(E_R\),
\(\theta\), or \(\kappa\)) would produce asymmetric payoff matrices
requiring bargaining-theoretic extensions. The qualitative result
(cooperation as the unique welfare-maximizing equilibrium) is expected
to hold under asymmetry, but the quantitative thresholds would depend on
the specific asymmetry structure.

\subsection{Applications}\label{applications}

The energy-denominated game theory framework applies to domains where
interactions involve measurable resource costs and repeated engagement:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Interstate conflict and arms races.} The friction multiplier
  \(\Phi\) provides a formal counterpart to the economic cost of
  military conflict between nations, and the cooperation threshold gives
  a quantitative criterion for when the threat of mutual defection
  (deterrence) suffices to sustain a peaceful equilibrium.
\item
  \textbf{Market competition and oligopoly.} Firms competing for market
  share face coordination costs (trade agreements, regulatory
  compliance) and conflict costs (price wars, patent litigation). The
  model predicts that industries with high friction costs
  (capital-intensive sectors, regulated markets) should exhibit more
  stable cooperative behavior (cartels, industry standards) than
  low-friction industries.
\item
  \textbf{Biological cooperation and altruism.} The cooperative
  equilibrium results extend naturally to weighted payoff functions that
  model inclusive fitness, coupled-agent systems, and belief-extended
  discount factors. These applications, which unify Hamilton's rule,
  reciprocal altruism, and group selection under a single
  payoff-expansion framework, are developed in \citep{Ethics-I}.
\end{enumerate}

\subsection{Future Work}\label{future-work}

\textbf{Continuous strategy spaces.} Extending from binary \(\{C, D\}\)
to continuous compliance \(\alpha \in [0,1]\) with energy-denominated
payoffs, to characterize the full cooperation manifold and verify that
partial defection is dominated.

\textbf{Imperfect monitoring.} Incorporating noisy observation channels,
where defection is detected with probability \(p < 1\), to bound
\(\delta^*_{\text{noisy}}\) as a function of detection probability and
determine when the baseline headroom (\(\delta^* = 0.363\) versus the
canonical \(0.5\)) is sufficient.

\textbf{Heterogeneous agents.} Agents with asymmetric endowments, time
preferences, or friction parameters, to determine whether the welfare
theorem generalizes to bargaining solutions and whether the equilibrium
selection result survives asymmetry.

\textbf{Coalition formation.} Modeling intermediate structures between
pairwise interaction and universal cooperation, to characterize when and
how stable coalitions form and how they resist internal defection.

\textbf{Embeddedness-dependent cooperation threshold.} The cooperation
threshold \(\delta^* \approx 0.363\) is derived for agents in a
shared-pool environment. Embedding agents in a production network, where
each agent's output feeds other agents' inputs, would lower this
threshold for agents with high in-degree and out-degree, because
defection would additionally sever supply lines that downstream agents
must rebuild at thermodynamic cost. Such an extension would provide a
first-principles derivation of why specialized, division-of-labor
economies exhibit more stable cooperation than autarkic ones.

\section{Conclusion}\label{conclusion}

We have proved that in the repeated multi-agent game with
energy-denominated payoffs, universal cooperation is the unique Nash
Equilibrium that simultaneously satisfies Pareto efficiency and welfare
maximization, for agents with \(\delta \geq \delta^*\)
(Theorem~\ref{thm-cooperation-theorem},
Theorem~\ref{thm-cooperation-maximizes-welfare},
Corollary~\ref{cor-cooperation-unique-efficient-ne}). The result rests
on the structural properties of the payoff matrix, derived from
thermodynamic parameters rather than stipulated, which make mutual
defection net-negative and cooperation self-reinforcing through
endogenous friction dynamics.

The cooperation threshold is substantially below the canonical \(0.5\)
of the standard Prisoner's Dilemma, and drops further to
\(\tilde{\delta}^* \approx 0.165\) when network friction is included
(Theorem~\ref{thm-network-adjusted-cooperation}). Defection is
evolutionarily unstable (Theorem~\ref{thm-defection-invasion-barrier}),
self-extinguishing (Corollary~\ref{cor-defection-self-extinction}), and
unprofitable even from a single deviation
(Theorem~\ref{thm-defection-profit-erasure}). These results hold for
arbitrarily large populations
(Corollary~\ref{cor-cooperation-strengthens-scale}).

\emph{Necessary Constraints} \citep{TC-I}, \emph{Strategic Entropy
Injection} \citep{TC-II}, and \emph{Thermodynamic Friction}
\citep{TC-IV} provide the physical infrastructure from which the payoff
structure derives, and \emph{Value Dynamics} \citep{TC-VI} analyzes the
dynamical stability of the resulting cooperative configurations.

\addcontentsline{toc}{section}{References}
\bibliography{../../shared/main}
\end{document}
